### 2020-07-30 paper/sex_selection_spacing-ver4.tex
Major editing of birth interval section(s) plus other edits
- remove delay effect discussion
- comment out conditional survival curves
- Reorganize birth interval discussion
- Remove average discussion of Estimation Strategy section

If parents realize that longer spacing leads to better child outcomes, we may even observe 
later conception, especially when, with sex selection, the next child is likely to be a boy.
Working in the opposite direction of this "delay" effect is that women with more education 
can, in principle, space their children closer together without substantially increasing 
child mortality risk. 
However, given the apparent lack of pressure to return to the labor force, this effect is 
likely small.

@@

For birth intervals, I estimate the average birth interval and a set of percentile
birth intervals.
To find the average birth interval, I calculate, for each woman, the probability of 
giving birth in each $t$, and her expected spell length from these probabilities.
I then average the individual expected spell lengths across women using their parity 
progression probabilities as weights.
Finally, I add nine months because spells begin nine months after the previous birth.
The process is similar for the percentile birth intervals, except that I use the time 
point when there is a given percent probability that a woman will give birth, conditional 
on her giving birth in the spell.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant changes, 
especially the spread of sex selection, that India experienced. 
I estimate the competing risk hazard model for each spell, education, and period 
combination and then predict average birth interval in months, sex ratio, and the 
probability of having a birth in that spell.

Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the predicted outcomes for the four education levels by area of residence.%
\footnote{
For legibility, the graphs do not show standard errors.
The graphed values, with standard errors, are shown in Appendix Tables 
\ref{tab:avg_sex_ratio_low} through \ref{tab:avg_sex_ratio_highest}.
Furthermore, I do not show the results for the fourth spell the best-educated women since 
the low number of births, especially with one or more boys, make the spacing and sex ratio 
results very noisy.
Results are available upon request.

[Urban low and rural highest in online Appendix]

}
For comparison, the graphs for the sex ratio also show the natural sex ratio, 
approximately 51.2 percent \citep{ben-porath76b,jacobsen99,Portner2015b}.

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_low_urban}
% \caption{combined graphs - low urban}
% \label{fig:spacing_low_urban}
% \end{figure}

@@

@@

@@

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_highest_rural}
% \caption{combined graphs - highest rural}
% \label{fig:spacing_highest_rural}
% \end{figure}

Average birth intervals have increased over time across all groups, but
the higher the parity and the higher the schooling, the larger the
increases.

In the latest period, the shortest average birth interval is 33 months for
women with no education in the second spell, an increase of a couple of
months from the first period.
The longest average interval was 51 months for the third spell for urban women with twelve 
or more years of education who had no boys, an increase of approximately a year over
4 decades.
For comparison, the median birth intervals reported for the NFHS have barely
moved over time, staying at approximately 31 months, and fast changing countries have 
median birth intervals that increase change around 3.5 months per decade 
\citep{Rutstein2011,Casterline2016}.%
\footnote{
Median estimates are in Appendix Tables \ref{tab:p25_p50_p75_low} through 
\ref{tab:p25_p50_p75_highest}.
}
This lack of change in the NFHS reported birth intervals over round underscores the 
importance of accounting for censoring of birth spells when trying to understand how birth 
spacing is changing over time.

Parity progression probabilities are lower the higher the education and are
falling across all groups over time, but the extent of the decline depends crucially on 
the sex composition of prior children and the parity.
Women with no sons are still the most likely to 
have the next birth, and this group has seen substantially smaller declines in
the likelihood of having another birth than women with one or more sons.%
\footnote{
However, there is also some evidence that parents prefer a mix of boys and girls: for 
the fourth spell women with two boys \emph{and} a girl
consistently have the lowest parity progression probability.
}
For urban women with eight to eleven years of education, for example, the second spell
probability of a next child declined by eight percentage points if the first child was a 
girl, but 15 percentage points if it was a boy.
Even more extreme is the fourth spell, where the likelihood of a birth declined by six 
percentage points to 75 percent if the woman had three children, but by almost 40 
percentage points to 25 percent if she had two boys and a girl.

\subsection{Distribution of Birth Spacing}

To better understand what factors are behind the changes in average spacing, I next 
examine its distribution---represented by the 25th and 75th percentiles 
of predicted birth intervals. 
This approach can capture differences across groups with similar changes in average spacing. 
For example, imagine that sex selection is behind an increase in average birth spacing, 
but that women still try to conceive early. 
In that case, we should see little change in the 25th percentile, but the 75th percentile 
spacing becomes much longer because those who use sex selection will have two or more 
pregnancies rather than one. 
In contrast, a general increase in spacing would show up as lengthening of both percentiles.

Figures \ref{fig:combined_p25_p75_urban} and \ref{fig:combined_p25_p75_rural} 
show how the predicted 25th and 75th percentile birth intervals have changed over time 
by spell and education.%
\footnote{
The Appendix shows conditional survival curves for all groups, and 
Appendix Tables \ref{tab:p25_p50_p75_low} through \ref{tab:p25_p50_p75_highest} 
show 25th, 50th, and 75th percentiles durations with bootstrapped standard errors.
}

For women with no or little education, the increases in average spacing
come predominantly from increases in the long end of the spacing
distribution rather than from a general increase.
Hence, spacing is still very short for many women, with 25 percent or more having
their next birth within two years of the prior birth, even in the last period.
Furthermore, the increases in the 75th percentile birth intervals are relatively small
for the second spell and slightly more substantial for the third spell at around five months.
Most of the divergence across sex compositions happen for the fourth spell, where 
the most significant increases are for women with two or three sons.
However, even for women with no or one son, the birth intervals have
increased by a half to a full year, consistent with some use of sex
selection.

The 25th percentile birth intervals for women with eight to eleven years of education also 
changed relatively little over time and is, as for the less-educated women, close to
around two years since the prior birth.
Hence, the increases in the longer end of the distribution drive
the change in average spacing.
For example, the 75th percentile spacing length for women with two girls in the third 
spell increased by more than 15 months, mainly as the result of more widespread use of sex 
selection as the sex ratios above indicate.

The most educated urban women show an unusual pattern for the third spell, where there 
is virtually no change in either the 25th or 75th percentile birth intervals in the
presence of at least one boy.
The 75th percentile birth interval, however, increases by more than 20 months,
corresponding to the massive increase in the sex ratio to almost 80 percent in the 
latest period.
Hence, the increases in the length of the longer intervals over time are
the main reason for the increases in average birth spacing and the
reversal in the spacing pattern by sex composition, consistent with
substantial use of sex selection.

Although most of the increases come from among the longer intervals, there is some 
evidence for the ``delay'' effect discussed above, where parents conceive later
precisely because they know that the next child is likely to be a boy since they plan
to use sex selection.
The best example is for the best-educated urban women in the third spell, where, in the 
last two periods, the spacing to the third child is significantly longer if the two prior 
children were girls than if they already had one or two boys.
These two periods are the periods where the predicted sex ratios are 70 and 80 percent,
respectively.
There is also a significantly longer spacing with only girls for women with eight to
eleven years of education in the last period for the third and fourth spell, although
the difference is substantially smaller.

\subsection{Sex Selection and Birth Spacing}

The results above show that average birth intervals increased over time across all groups 
and that these increases are larger for higher parities and with more schooling. 
Furthermore, the increases come predominately from the longer intervals getting even 
longer while there is little change in the length of the shortest intervals.

What do these results tell us about the extent to which the introduction of sex selection 
changed birth spacing across sex compositions? 
There are three broad groups in the data.

The first group consists of women who do not use sex selection even if they have only 
girls, such as rural women with no education. 
For this group, the shortest spacing---across all spells---is when they have only girls. 
Notably, the difference in birth intervals across sex compositions has even grown over 
time as spacing when sons are present increased.

In the second group are women who use sex selection and who, therefore, show the expected 
associated increase in birth intervals for those sex compositions. 
However, the intervals for their other sex compositions have increased even more. 
The result is that the relative pattern remains more or less the same. 
An example is women with eight more years of education in the second spell.

The final group consists of those women, who, in the absence of a son, have the most 
biased sex ratios and the longest birth intervals, and where this spacing is longer than 
for the other possible sex compositions. 
The most extreme example is best-educated urban women with two girls, who have a sex ratio 
approaching 80 and an average birth interval of 51 months. 
This birth interval is seven months longer than with at least one boy.

Even more striking is the change for this group when sex selection was first introduced.
There was a six months jump, from 39.5 to 45.4, in average spacing between the first and 
the second period for women with two girls, but only a month increase for the other sex 
compositions.
Furthermore, the 75th percentile birth interval with two girls increased from 48 months to 
64 months with two girls over the period, while the other sex compositions showed a 
slightly decrease from around 55 months to 54 months.

Theses changes may even be an underestimate of the changes in birth spacing sine
this particular group appear to have had access to sex selection even before it became
widespread as shown by the unequal sex ratio for the 1972--1984 period for women with
two girls.

In summary, although birth spacing has increased over time with improving economic 
conditions and falling fertility, most of the increases in birth spacing come from the
increased use of sex selection. 
The most substantial increase over the four decades is a full year for the best-educated 
women with two girls, which is the group with the most biased sex ratio.
There is no clear evidence for or against a reversal in the use of sex selection, with 
some showing increases in sex ratios between the last two periods, others little change, 
and some a decline.

@@

Despite the longer average birth intervals, a large number of women still have very 
short birth intervals. 
Hence, the changes in birth spacing are driven predominately by the longer intervals 
becoming even longer over time.
This pattern holds generally, but is especially prevalent for those who use sex selection 
and, therefore, may end up with multiple abortions. 
The exception to the small change in short spacing is the delay in conception for women 
with very high use of sex selection, presumably because this group knows that
the next child will be a boy.

@@

\begin{figure}
\centering
\subfloat[NFHS-1]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_1}}
\subfloat[NFHS-2]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_2}} \\
\subfloat[NFHS-3]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_3}} 
\subfloat[NFHS-3]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_4}} 
\caption{Ratio of Boys for ``First'' Births by Survey Round}
\label{fig:sex_ratio_recall_rounds_bo1}
\end{figure}

\begin{figure}
\centering
\subfloat[NFHS-1]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_1_bo2}}
\subfloat[NFHS-2]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_2_bo2}} \\
\subfloat[NFHS-3]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_3_bo2}} 
\subfloat[NFHS-3]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_4_bo2}} 
\caption{Ratio of Boys for ``Second'' Births by Survey Round}
\label{fig:sex_ratio_recall_rounds_bo2}
\end{figure}

@@

\section{Duration Results Tables}

@@

The first set of tables, Tables \ref{tab:avg_sex_ratio_low}, \ref{tab:avg_sex_ratio_med}, 
\ref{tab:avg_sex_ratio_high}, and \ref{tab:avg_sex_ratio_highest}, show predicted average 
birth intervals, sex ratios, and 
probabilities of having a birth by decade, spell, and sex composition for the three 
education levels separated by the area of residence, together with standard errors for all 
three outcomes.
The standard errors are based on bootstrapping for all three measures,
where the model is repeatedly estimated using resampling with replacement.

@@

\begin{figure}[htpb]
\captionsetup[subfigure]{labelformat=empty,position=top}
\captionsetup{font=footnotesize,skip=2pt}
\centering
\caption*{No education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_low_urban}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_low_urban}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_low_urban}}
\\
\caption*{One to Seven Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_med_urban}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_med_urban}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_med_urban}}
\\
\caption*{Eight to Eleven Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_high_urban}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_high_urban}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_high_urban}}
\\
\caption*{Twelve or More Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_highest_urban}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_highest_urban}}
\begin{minipage}{0.32\textwidth}\hspace{1cm}\end{minipage}
\captionsetup{font=normalsize}
\caption{
Changes in 25th Percentile Birth Intervals for Urban Women by Spell and Education
}
\label{fig:p25_urban}
\end{figure}

\begin{figure}[htpb]
\captionsetup[subfigure]{labelformat=empty,position=top}
\captionsetup{font=footnotesize,skip=2pt}
\centering
\caption*{No education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_low_rural}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_low_rural}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_low_rural}}
\\
\caption*{One to Seven Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_med_rural}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_med_rural}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_med_rural}}
\\
\caption*{Eight to Eleven Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_high_rural}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_high_rural}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_high_rural}}
\\
\caption*{Twelve or More Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_highest_rural}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_highest_rural}}
\begin{minipage}{0.32\textwidth}\hspace{1cm}\end{minipage}
\captionsetup{font=normalsize}
\caption{
Changes in 25th Percentile Birth Intervals for Rural Women by Spell and Education
}
\label{fig:p25_rural}
\end{figure}

\begin{figure}[htpb]
\captionsetup[subfigure]{labelformat=empty,position=top}
\captionsetup{font=footnotesize,skip=2pt}
\centering
\caption*{No education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p50_spell2_low_urban}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p50_spell3_low_urban}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p50_spell4_low_urban}}
\\
\caption*{One to Seven Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p50_spell2_med_urban}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p50_spell3_med_urban}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p50_spell4_med_urban}}
\\
\caption*{Eight to Eleven Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p50_spell2_high_urban}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p50_spell3_high_urban}}
\subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p50_spell4_high_urban}}
\\
\caption*{Twelve or More Years of Education}
\subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p50_spell2_highest_urban}}
\subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p50_spell3_highest_urban}}
\begin{minipage}{0.32\textwidth}\hspace{1cm}\end{minipage}
\captionsetup{font=normalsize}
\caption{
Changes in 50th Percentile Birth Intervals for Urban Women by Spell and Education
}
\label{fig:p50_urban}
\end{figure}

>>>>

% If parents realize that longer spacing leads to better child outcomes, we may even observe 
% later conception, especially when, with sex selection, the next child is likely to be a boy.
% Working in the opposite direction of this "delay" effect is that women with more education 
% can, in principle, space their children closer together without substantially increasing 
% child mortality risk. 
% However, given the apparent lack of pressure to return to the labor force, this effect is 
% likely small.

@@

For birth intervals, I estimate a set of percentile birth intervals.
To find a given percentile birth interval, I first calculate for each woman when there is 
a given percent chance that she will have given birth, conditional on the probability of 
giving birth in the spell.
I then add nine months to account for the start of the spell. 
For example, with an 80 percent chance of giving birth by the end of the spell, the median birth 
interval is the predicted number of months before a woman passes the 40 percent mark on her 
survival curve. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant changes, 
especially the spread of sex selection, that India experienced. 
I estimate the competing risk hazard model for each spell, education, and period 
combination and then predict 25th, 50th, and 75th percentiles birth intervals in 
months, the sex ratio, and the probability of having a birth in that spell.
Sex selection increases all three percentile birth intervals, but the effects of abortions 
are easiest to observe in the 75th percentile birth interval, 
and it is also the changes in the longest birth intervals that will likely the largest 
impact on fertility measures.
The 25th percentile captures changes in the short spacing, which is the most likely to 
have an adverse impact on health.

Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the predicted outcomes for the four education levels by area of residence.
The underlying values with bootstrapped standard errors are available in the online 
appendix.
The sex ratio graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.
Results for urban women without education, rural women with twelve or more years of 
education, and the fourth spell for women with twelve or more years of education are
presented in the online appendix because they are based on relatively few observations.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_low_urban}
\caption{combined graphs - low urban}
\label{fig:spacing_low_urban}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_highest_rural}
\caption{combined graphs - highest rural}
\label{fig:spacing_highest_rural}
\end{figure}

@@

The results show three broad trends.
% [Birth intervals]
First, 
over the four decades, birth intervals have lengthen, whether measured by the
25th, 50th, or 75th percentile.
These increases are larger the higher the education of the mother, the higher the parity, 
and the higher the percentile measure.

% [Parity progression probability]
Second, 
the likelihood of a next birth has fallen over time, and the declines are
larger the higher the education, the higher the parity, and with the presence of at
least one son. 
Within a given spell and period, parity progressions are lower in urban than in rural areas, 
if at least one son is present, and the more educated the mother.

% [Sex ratios]
Finally, 
the sex ratios of next births has become more male dominated for women with no
sons, while there are no clear trends for the other sex compositions.
The increases in the predicted percent boys are larger the higher the education and the 
higher the parity.
Furthermore, for a given spell and period, sex ratios in the absence of a son are higher 
in urban than in rural areas, the higher the education of the mother, and the higher
the parity.
There is some evidence of unnaturally high percentage boys born even when women have one 
son, although most are clearly lower than in the absence of a son and the failing fertility
make more of those estimate noisy.

% [25th percentile birth intervals]
Despite the increases in birth intervals and the decline in fertility, a remarkably high 
proportion of birth intervals are still very short.
For all but the most educated, more than 25 percent of women have their second and their
third child within 24 months of their previous child.
Furthermore, even with the larger increases in birth intervals with higher parity, the 
25th percentile birth intervals for the fourth spell is around 24 months for women with 
less than eight years of education.

% [median birth intervals]
The median birth intervals have increased more than the 25th percentile birth intervals
but are still also relatively short.
Although most median birth intervals have increased by three to six months over the four 
decades, the majority of median birth intervals are still at 36 months or below, 
with the shortest only 29 months.
For comparison, fast changing countries have median birth intervals that increase around 
3.5 months per decade with declining fertility \citep{Rutstein2011,Casterline2016}.
Furthermore, the NFHS reports show median closed birth intervals that have barely
moved over time, staying at approximately 31 months, underscoring the importance of 
accounting for censoring of birth spells when trying to understand how birth spacing 
changes over time.

An important exception to the relatively short median intervals come from the better 
educated women, especially in the absence of sons.
The longest median interval was 48 months for the third spell for urban women with twelve 
or more years of education who had no boys, an increase of almost 15 months over
four decades.
Similary, urban women with eight to eleven years of education and no sons had median
birth intervals of around 40 months, an increase of about 10 months.
Finally, the best educated women now have a median second birth interval of around 42
months whether their first child was a boy or a girl.

% [75th percentile birth intervals]
The most substantial differences across women come from the longer birth intervals, 
captured by the 75th percentile birth intervals.
For the two lowest education groups the increases in the 75th percentile birth 
intervals are relatively small for the second spell and slightly more substantial for the 
third spell at around five months.
However, the fourth spell show significant increases, especially for women with two
or three sons, with increases of fifteen to twenty months.
The fourth spell is also where most of the divergence across sex compositions happen,
but even for women with no or one son, the 75th percentile birth intervals have
increased by a half to a full year.
For women with eight to eleven years of education, the 75th percentile spacing length for 
women with two girls in the third spell increased by more than 15 months, mainly as the 
result of more widespread use of sex selection as the sex ratios above indicate.

% [Exceptions to the results]
The best-educated women are an especially interesting case that run somewhat counter
to the results above.
First, despite a rapidly declining probability of having a third child when they already 
have a son, there is little change in the associated birth intervals and all are lower for 
the third spell than for the second.
Second, the massive increase in the sex ratio to almost 80 percent boys for women with two 
girls is associated with a more than 20 month increase in the 75th percentile birth 
interval.
The third spell is also the case with the clearest reversal in the spacing pattern by sex 
composition; the birth intervals with two girls are consistently longer than with one
or two sons no matter the percentile used.

What do these results tell us about the role of son preference and the extent to which 
the introduction of sex selection changed birth spacing across sex compositions?

The clearest illustration of how the combination of son preference and the introduction of 
sex selection impacted birth spacing comes from the third spell of the best-educated urban 
women.
Women with two girls have a predicted sex ratio for their third child approaching 80 
percent boys and a 75th percentile birth interval of almost 70 months, about 13 months
longer than if they had at least one son.
Even more striking is the change for this group when sex selection was first introduced.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months with two girls, while the other sex compositions showed a 
slightly decrease from around 55 months to 54 months.
Theses changes may even be an underestimate of the changes in birth spacing sine
this particular group appear to have had access to sex selection even before it became
widespread as shown by the unequal sex ratio for the 1972--1984 period for women with
two girls.

Even for the women the least likely to use sex selection, such as rural women with no 
education, son preference shows up clearly. 
For this group, the shortest spacing---across all spells---is when they have only girls. 
Notably, the difference in birth intervals across sex compositions has grown over 
time as spacing when sons are present increased.
Despite the strong son preference in spacing and sex ratios, there is also some indication 
that parents prefer a mix of boys and girls;
women with two boys \emph{and} a girl consistently have the lowest parity progression 
probability in the fourth spell across education levels.

Did the predictions of a turn-around in the use of sex selection come true?
There is no clear evidence for or against a reversal in the use of sex selection, with 
some showing increases in sex ratios between the last two periods, others little change, 
and some a decline.
The best-educated women is again a good illustration.
The percent boys born to women with two girls continued to increase over the last two
periods, but the likelihood of a third birth declined and when the first child was a girl, 
the sex ratio for the second birth dropped slightly as did the probability of a second birth. 
There is, however, also cases where there is clearly no abatement in the increasing use of 
sex selection. 
The sex ratios in the absence of girls continued to increase for rural women with
one to seven years of education and the likelihood of an additional birth remained high.

% [Interpretation]
How do these changes in birth intervals tie in with the changes in India over the four
decades?
% [TK economic growth? -> higher return to education for males -> lower fertility and
% higher investment in boys' education -- has anybody looked at this?]
First, the parity progression probabilities have been falling over time across all 
education levels, especially for women with at least one son.
Falling fertility is associated with increases in longer spacing, although it is still an 
unresolved question why \citep{Casterline2016}.
What is especially interesting is that the results for the most educated women show that 
the rapidly declining probability of a third birth for women with one or two sons did not 
lengthen the birth intervals at any of the percentiles.
One possibility is that the most educated women have both better access to modern
contraceptive and are better at using traditional contraceptive methods 
\citep{Rosenzweig1989}.

Second, there is evidence that in situations where sex selection does not play a role 
in birth intervals, women with the smallest change in birth intervals are also those most 
likely to work.
For example, the increases in average third spell birth intervals for urban women with
one boy and one girl over the four decades were 2.7, 3.4, 5.8, and 1.8 months from lowest 
to highest education groups.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0, respectively.
See the online appendix tables for the average birth intervals.
}
This means that the group with the lowest labor force participation, women with eight to 
eleven years of education, was also the group with the largest increases in average 
spacing, likely driven by the substantial improvement in household income that came
from economic growth.

Finally, although birth spacing has increased over time with improving economic 
conditions and falling fertility, most of the increases in birth spacing come from the
increased use of sex selection. 
The most substantial increase over the four decades is a full year for the best-educated 
women with two girls, which is the group with the most biased sex ratio.

@@

Despite the longer average birth intervals, a large number of women still have very 
short birth intervals. 
Hence, the changes in birth spacing are driven predominately by the longer intervals 
becoming even longer over time.
This pattern holds generally, but is especially prevalent for those who use sex selection 
and, therefore, may end up with multiple abortions.

@@

\begin{figure}
\centering
\subfloat[NFHS-1]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_1}}
\subfloat[NFHS-2]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_2}} \\
\subfloat[NFHS-3]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_3}} 
\subfloat[NFHS-4]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_4}} 
\caption{Ratio of Boys for ``First'' Births by Survey Round}
\label{fig:sex_ratio_recall_rounds_bo1}
\end{figure}

\begin{figure}
\centering
\subfloat[NFHS-1]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_1_bo2}}
\subfloat[NFHS-2]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_2_bo2}} \\
\subfloat[NFHS-3]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_3_bo2}} 
\subfloat[NFHS-4]{\includegraphics[width=.49\textwidth]{recall_sex_ratio_marriage_round_4_bo2}} 
\caption{Ratio of Boys for ``Second'' Births by Survey Round}
\label{fig:sex_ratio_recall_rounds_bo2}
\end{figure}

@@

\section{Additional Results Figures and  Tables}

@@

Figures \ref{fig:spacing_low_urban} and \ref{fig:spacing_highest_rural} show 25th, 50th,
and 75th percentile birth intervals, the sex ratio, and the probability of parity
progression by spell for urban women with no education and rural women with twelve or
more years of education, respectively.

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_low_urban}
\caption{combined graphs - low urban}
\label{fig:spacing_low_urban}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_highest_rural}
\caption{combined graphs - highest rural}
\label{fig:spacing_highest_rural}
\end{figure}

The first set of tables, Tables \ref{tab:p25_p50_p75_low}, \ref{tab:p25_p50_p75_med},
\ref{tab:p25_p50_p75_high}, and \ref{tab:p25_p50_p75_highest} show 25th, 50th, and
75th percentile birth intervals together with their standard errors.
The standard errors for all measures are based on bootstrapping,
where the model is repeatedly estimated using resampling with replacement.

The second set of tables, 
Tables \ref{tab:avg_sex_ratio_low}, \ref{tab:avg_sex_ratio_med}, 
\ref{tab:avg_sex_ratio_high}, and \ref{tab:avg_sex_ratio_highest}, show predicted average 
birth intervals, sex ratios, and probabilities of having a birth by decade, spell, and sex 
composition for the four education levels separated by the area of residence, together 
with bootstrapped standard errors for all three outcomes.
To find the average birth interval, I calculate, for each woman, the probability of 
giving birth in each t, and her expected spell length from these probabilities. 
I then average the individual expected spell lengths across women using their parity 
progression probabilities as weights. Finally, I add nine months because spells begin 
nine months after the previous birth.

@@

% \begin{figure}[htpb]
% \captionsetup[subfigure]{labelformat=empty,position=top}
% \captionsetup{font=footnotesize,skip=2pt}
% \centering
% \caption*{No education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_low_urban}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_low_urban}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_low_urban}}
% \\
% \caption*{One to Seven Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_med_urban}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_med_urban}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_med_urban}}
% \\
% \caption*{Eight to Eleven Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_high_urban}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_high_urban}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_high_urban}}
% \\
% \caption*{Twelve or More Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_highest_urban}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_highest_urban}}
% \begin{minipage}{0.32\textwidth}\hspace{1cm}\end{minipage}
% \captionsetup{font=normalsize}
% \caption{
% Changes in 25th Percentile Birth Intervals for Urban Women by Spell and Education
% }
% \label{fig:p25_urban}
% \end{figure}
% 
% 
% \begin{figure}[htpb]
% \captionsetup[subfigure]{labelformat=empty,position=top}
% \captionsetup{font=footnotesize,skip=2pt}
% \centering
% \caption*{No education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_low_rural}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_low_rural}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_low_rural}}
% \\
% \caption*{One to Seven Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_med_rural}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_med_rural}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_med_rural}}
% \\
% \caption*{Eight to Eleven Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_high_rural}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_high_rural}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p25_spell4_high_rural}}
% \\
% \caption*{Twelve or More Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p25_spell2_highest_rural}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p25_spell3_highest_rural}}
% \begin{minipage}{0.32\textwidth}\hspace{1cm}\end{minipage}
% \captionsetup{font=normalsize}
% \caption{
% Changes in 25th Percentile Birth Intervals for Rural Women by Spell and Education
% }
% \label{fig:p25_rural}
% \end{figure}
% 
% \begin{figure}[htpb]
% \captionsetup[subfigure]{labelformat=empty,position=top}
% \captionsetup{font=footnotesize,skip=2pt}
% \centering
% \caption*{No education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p50_spell2_low_urban}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p50_spell3_low_urban}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p50_spell4_low_urban}}
% \\
% \caption*{One to Seven Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p50_spell2_med_urban}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p50_spell3_med_urban}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p50_spell4_med_urban}}
% \\
% \caption*{Eight to Eleven Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p50_spell2_high_urban}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p50_spell3_high_urban}}
% \subfloat[][Fourth Spell]{\includegraphics[width=0.32\textwidth]{p50_spell4_high_urban}}
% \\
% \caption*{Twelve or More Years of Education}
% \subfloat[][Second Spell]{\includegraphics[width=0.32\textwidth]{p50_spell2_highest_urban}}
% \subfloat[][Third Spell]{\includegraphics[width=0.32\textwidth]{p50_spell3_highest_urban}}
% \begin{minipage}{0.32\textwidth}\hspace{1cm}\end{minipage}
% \captionsetup{font=normalsize}
% \caption{
